---
title: "Assignment 2"
author: "Marion"
date: "2025-05-11
output: html_document"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
install.packages("purrr")
library(purrr)
library(class)
library(caret)
library(ggplot2)
library(lattice)
library(MASS)
library(boot)
library(ggcorrplot)
library(corrplot)
library(lmtest)
library(tseries)
library(dplyr)
library(lubridate)
library(tidyr)
```

#Data loading

```{r}
weather<-read.csv("https://raw.githubusercontent.com/Marion196cmd/File-1/refs/heads/main/Summary%20of%20Weather.csv")
station<-read.csv("https://raw.githubusercontent.com/Marion196cmd/File-1/refs/heads/main/Weather%20Station%20Locations.csv")
```
#Let us now edit the data
```{r}
combined_data <- left_join(weather, station, by = c("STA" = "WBAN"))
data <- combined_data %>% select(where(~ !any(is.na(.))))
head(data)
```
#Let us first visualise some of our data 
```{r}
data$Snowfall[data$snowfall %in% "NA"] <- 0
data$PRCP[data$PRCP %in% "NA"] <- 0
data$snowfall<-as.numeric(data$Snowfall)
data$PRCP<-as.numeric((data$PRCP))
elev_group<-cut(data$ELEV, breaks = c(0, 500, 1000, 1500, 2000, Inf))
log_PRCP <-log(data$PRCP + 0.001)
temp_diff<- data$MaxTemp - data$MinTemp
# Temperature distribution
ggplot(data, aes(x=MeanTemp)) +
  geom_histogram(fill="skyblue", bins=30) +
  ggtitle("Distribution of Mean Temperatures")
# Geography interactions
ggplot(data, aes(x=Latitude, y=MeanTemp, color=elev_group)) +
  geom_point(alpha=0.5) +
  geom_smooth(method="lm") +
  ggtitle("Temperature vs Latitude by Elevation group")

# Precipitation effects
ggplot(data, aes(x=log_PRCP, y=MeanTemp)) +
  geom_point(aes(color=Latitude)) +
  geom_smooth(method="loess", color="red") +
  facet_wrap(~ elev_group) +
  ggtitle("Temperature vs Log Precipitation by Elevation group")
```
#Let us create a model accounting temperature
```{r}
wwii_weather_clean<- data%>%
mutate(
    Date = as.Date(Date, format="%Y-%m-%d"), 
    Precip = as.numeric(ifelse(Precip == "T", 0.001, Precip)), 
    PRCP = as.numeric(ifelse(PRCP == "T", 0.001, PRCP)), 
    SNF = as.numeric(ifelse(SNF == "T", 0.001, SNF)),
    DSN = as.numeric(Date - as.Date(paste0(year(Date), "-11-01"))), 
    DSN = ifelse(DSN < 0, 365 + DSN, DSN)
  ) %>%
  filter(complete.cases(.))
ggplot(wwii_weather_clean, aes(x=MaxTemp, y=MinTemp)) +
  geom_point() +
  geom_smooth(method="lm")
model <- lm(MeanTemp ~ MaxTemp + MinTemp + PRCP+ DSN+ I(DSN^2), data=wwii_weather_clean)
summary(model)
```
#In order to detect multicolinearity allow us to compute a correlation matrix
```{r}
cor_data <- wwii_weather_clean %>%
  mutate(
    MaxTemp = as.numeric(MaxTemp),
    MinTemp = as.numeric(MinTemp),
    PRCP = as.numeric(PRCP),  
    Snowfall = as.numeric(Snowfall)  
  ) %>%
  select(MaxTemp, MinTemp, PRCP, Snowfall)
cor_matrix <- cor(cor_data, use = "complete.obs")  
corrplot::corrplot(cor_matrix, method = "circle")

```

#Let us now do a block bootstrap method and fit our model 
```{r}
set.seed(123)
n <- nrow(wwii_weather_clean)
L <- 7 
B <- 1000
block_bootstrap <- function(data, model_formula, L, B) {
  coef_matrix <- matrix(NA, nrow=B, ncol=length(coef(model)))
  pred_matrix <- matrix(NA, nrow=B, ncol=n)
  for(i in 1:B) {
     block_starts <- sample(1:(n-L+1), size=ceiling(n/L), replace=TRUE)
    idx <- unlist(purrr::map(block_starts, ~.x:(.x+L-1))) %>% 
      head(n) %>% 
      na.omit()
    boot_data <- data[idx, ]
    boot_model <- lm(formula(model_formula), data=boot_data)
    coef_matrix[i, ] <- coef(boot_model)
    pred_matrix[i, ] <- predict(boot_model, newdata=data)
  }
  return(list(coefs=coef_matrix, preds=pred_matrix))
}
boot_results <- block_bootstrap(wwii_weather_clean, 
                               formula(model), L, B)
coef_ci <- apply(boot_results$coefs, 2, quantile, probs=c(0.025, 0.975))
colnames(coef_ci) <- names(coef(model))
```
#We will compare the predicted values and find model metrics
```{r}
pred_ci <- apply(boot_results$preds, 2, quantile, probs=c(0.025, 0.975))

wwii_weather_clean <- wwii_weather_clean %>%
  mutate(
    pred_mean = colMeans(boot_results$preds),
    pred_lower = pred_ci[1, ],
    pred_upper = pred_ci[2, ]
  )
rmse <- sqrt(mean((wwii_weather_clean$MeanTemp - wwii_weather_clean$pred_mean)^2))
ggplot(wwii_weather_clean, aes(x=Date)) +
  geom_line(aes(y=MeanTemp), color="black") +
  geom_ribbon(aes(ymin=pred_lower, ymax=pred_upper), fill="blue", alpha=0.3) +
  labs(title="Temperature Predictions with 95% CI", y="Mean Temperature")
par(mfrow=c(2,2))
plot(model)
```

 # Fit a model also considering geographical factors.
```{r}
data$Date <- as.Date(data$Date)
data <- data %>%
  filter(
    !is.na(ELEV),
    !is.na(Date),
    !is.na(MeanTemp),
    !is.na(MaxTemp),
    !is.na(MinTemp),
    !is.na(PRCP),
    !is.na(Latitude)
  )
data <- data %>%
  mutate(
    elev_group = cut(
      ELEV,
      breaks = c(0, 500, 1000, 1500, 2000, max(ELEV, na.rm = TRUE)),
      include.lowest = TRUE
    ),
    log_PRCP = log(PRCP + 0.001),
    DSN = as.numeric(Date - as.Date(paste0(year(Date), "-11-01"))),
    DSN = ifelse(DSN < 0, 365 + DSN, DSN)
  )
modeltotal <- lm(
  MeanTemp ~ MaxTemp + MinTemp + log_PRCP * elev_group + log_PRCP * Latitude + DSN + I(DSN^2),
  data = data
)
summary(modeltotal)

```
#Let us update our block bootstrap to fit this new model
```{r}
set.seed(123)
newblock_bootstrap <- function(data, model_formula, L, B) {
  full_model <- lm(model_formula, data = data)
  coef_names <- names(coef(full_model))
  n <- nrow(data)
  coef_matrix <- matrix(NA, nrow = B, ncol = length(coef_names))
  colnames(coef_matrix) <- coef_names
  pred_matrix <- matrix(NA, nrow = B, ncol = n)
  for (i in 1:B) {
    block_starts <- sample(1:(n - L + 1), size = ceiling(n / L), replace = TRUE)
    idx <- unlist(purrr::map(block_starts, ~ .x:(.x + L - 1))) %>%
      head(n) %>%
      na.omit()
    boot_data <- data[idx, ]
    boot_data$elev_group <- factor(boot_data$elev_group, levels = levels(data$elev_group))
    boot_model <- tryCatch(
      lm(model_formula, data = boot_data),
      error = function(e) return(NULL)
    )
    if (!is.null(boot_model)) {
      coefs <- coef(boot_model)
      coef_matrix[i, names(coefs)] <- coefs
      pred_matrix[i, ] <- tryCatch(
        predict(boot_model, newdata = data),
        error = function(e) rep(NA, n)
      )
    }
  }
  return(list(coefs = coef_matrix, preds = pred_matrix))
}
newboot<- newblock_bootstrap(data, formula(modeltotal), L, B)
```

#let us now look at the new model metrics
```{r}
newpred_ci <- apply(newboot$preds, 2, quantile, probs = c(0.025, 0.975))
data <- data %>%
  mutate(
    newpred_mean = colMeans(newboot$preds),
    newpred_lower = newpred_ci[1, ],
    newpred_upper = newpred_ci[2, ]
  )
rmse <- sqrt(mean((data$MeanTemp - data$newpred_mean)^2))
ggplot(data, aes(x = Date)) +
  geom_line(aes(y = MeanTemp), color = "black") +
  geom_line(aes(y = newpred_mean), color = "blue") +
  geom_ribbon(aes(ymin = newpred_lower, ymax = newpred_upper), fill = "lightblue", alpha = 0.4) +
  labs(
    title = "Mean Temperature: Observed vs Predicted (New model)",
    y = "Mean Temperature (°C)"
  ) +
  theme_minimal()
par(mfrow=c(2,2))
plot(modeltotal)
```

#Let us now compare the 2 models by looking 
```{r}
install.packages(purrr)
library(purrr)
dwtest(model)
dwtest(modeltotal)
acf(residuals(model), main="Residual autocorellation of simple model")
acf(residuals(modeltotal), main="Residual autocorellation of updated model")
models <- list(
  "Base" = model,
  "Final"= modeltotal
)
model_metrics <- map_df(models, ~{
  data.frame(
    AIC = AIC(.x),
    BIC = BIC(.x),
    R2_adj = summary(.x)$adj.r.squared,
    RMSE = sqrt(mean(residuals(.x)^2))
  )
}, .id="Model")
knitr::kable(model_metrics, caption="Model Comparison Metrics")
```
# Select a target station (example: Station 381 with specific geography)
```{r}
target_station <- data %>%
  filter(STA == 381) %>%
  slice(1) %>%
  select(Latitude, ELEV, elev_group) %>%
  distinct()
create_scenarios <- function(original_data) {
  modified_data <- original_data %>%
    filter(STA == 381) %>%
    mutate(
      PRCP = PRCP * 1.10,
      log_PRCP = log(PRCP + 0.001),
      MaxTemp = MaxTemp + 2,
      MinTemp = MinTemp + 2,
      MeanTemp = MeanTemp + 2,
      elev_PRCP = ELEV * log_PRCP,
      lat_PRCP = Latitude * log_PRCP
    )
  bind_rows(
    original_data %>%
      mutate(scenario = "baseline"),
    modified_data %>%
      mutate(scenario = "modified")
  )
}
scenario_predictions <- create_scenarios(data) %>%
  group_by(scenario) %>%
  mutate(
    pred = predict(modeltotal, newdata = cur_data()),
    pred_diff = pred - mean(pred, na.rm = TRUE)
  )
scenario_bootstrap <- function(data, model_formula, L = 7, B = 1000, modification_fn = NULL) {
  full_model <- lm(model_formula, data = data)
  coef_names <- names(coef(full_model))
  n <- nrow(data)
  coef_matrix <- matrix(NA, nrow = B, ncol = length(coef_names))
  colnames(coef_matrix) <- coef_names
  pred_matrix <- matrix(NA, nrow = B, ncol = n)
  for (i in 1:B) {
    block_starts <- sample(1:(n - L + 1), ceiling(n / L), replace = TRUE)
    idx <- unlist(purrr::map(block_starts, ~.x:(.x + L - 1))) %>%
      head(n) %>%
      na.omit()
    boot_data <- if (!is.null(modification_fn)) {
      modification_fn(data[idx, ])
    } else {
      data[idx, ]
    }
    boot_data$elev_group <- factor(boot_data$elev_group, levels = levels(data$elev_group))
    boot_model <- tryCatch(
      lm(model_formula, data = boot_data),
      error = function(e) NULL
    )
    
    if (!is.null(boot_model)) {
      coefs <- coef(boot_model)
      coef_matrix[i, names(coefs)] <- coefs
      pred_matrix[i, ] <- tryCatch(
        predict(boot_model, newdata = data),
        error = function(e) rep(NA, n)
      )
    }
  }
  
  return(list(coefs = coef_matrix, preds = pred_matrix))
}
modify_station <- function(data) {
  data %>%
    mutate(
      PRCP = ifelse(ELEV > 500, PRCP * 1.10, PRCP),
      log_PRCP = log(PRCP + 0.001),
      elev_PRCP = ELEV * log_PRCP,
      lat_PRCP = Latitude * log_PRCP
    )
}
set.seed(3125)
L <- 7
B <- 1000
boot_orig <- newblock_bootstrap(data, formula(modeltotal), L, B)
boot_mod <- scenario_bootstrap(data, formula(modeltotal), L, B, modify_station)
scenario_comparison <- bind_rows(
  tibble(
    Scenario = "Original",
    RMSE = sqrt(mean((data$MeanTemp - colMeans(boot_orig$preds))^2)),
    Temp_Effect = mean(colMeans(boot_orig$coefs[, grep("log_PRCP", colnames(boot_orig$coefs))]))
  ),
  tibble(
    Scenario = "Modified",
    RMSE = sqrt(mean((data$MeanTemp - colMeans(boot_mod$preds))^2)),
    Temp_Effect = mean(colMeans(boot_mod$coefs[, grep("log_PRCP", colnames(boot_mod$coefs))]))
  )
)
ggplot(scenario_comparison, aes(x = Scenario, y = Temp_Effect, fill = Scenario)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = Temp_Effect * 0.9, ymax = Temp_Effect * 1.1), width = 0.2) +
  labs(
    title = "Temperature Sensitivity to Precipitation Changes",
    y = "Precipitation Effect Coefficient",
    subtitle = "With 10% increased precipitation at high elevation stations"
  ) +
  theme_minimal()
coef_comparison <- bind_rows(
  boot_orig$coefs %>% as_tibble() %>% mutate(Scenario = "Original"),
  boot_mod$coefs %>% as_tibble() %>% mutate(Scenario = "Modified")
) %>%
  pivot_longer(-Scenario, names_to = "Coefficient", values_to = "Value")
ggplot(coef_comparison, aes(x = Value, fill = Scenario)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~Coefficient, scales = "free") +
  labs(title = "Bootstrap Coefficient Distributions by Scenario") +
  theme_minimal()
```